---
title: "Real-time estimation of the difference in the test seeking distribution of Omicron compared to Delta in England using S-Gene Target Status as a Proxy"
subtitle: "Summary report"
author: Sam Abbott (1)
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
date: "`r format(Sys.Date(), format = '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    toc_depth: 4
    includes:
      before_body: header.html
      after_body:  footer.html
  pdf_document:
    extra_dependencies: ["float"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE,
                      warning = FALSE, message = FALSE,
                      root.dir = here::here(),
                      fig.width = 12, fig.height = 9)
options(digits = 1)
library(here)
library(forecast.vocs)
library(dplyr)
library(loo)
library(scoringutils)
library(knitr)
library(data.table)
```

```{r load-results}
# load packages
library(ggplot2)
library(scales)
library(dplyr)
library(here)
library(readr)
library(data.table)

# load functions
source(here("R", "load-local-data.R"))
source(here("R", "plot-summary.R"))
source(here("R", "plot-daily-cases.R"))
source(here("R", "plot-omicron-95.R"))
source(here("R", "plot-cumulative-percent.R"))

# get latest date
date_latest <- get_latest_date()

# load data
daily <- load_local_data(date_latest)
```

1. Centre for the Mathematical Modelling of Infectious Diseases, London School of Hygiene & Tropical Medicine, London WC1E 7HT, United Kingdom


## Introduction

## Methods

### Data

We use S-gene status by specimen date sourced from UKHSA as a direct proxy for the Omicron variant with a target failure indicating a case has the Omicron variant. Data was available by UKHSA region.

### Models

### Statistical Inference

### Implementation

All models were implemented using the [`forecast.vocs` R package](https://epiforecasts.io/forecast.vocs/) [@R; @forecast.vocs] and fit using `stan` [@stan] and `cmdstanr` [@cmdstanr]. Each model was fit using 2 chains with each chain having 1000 warmup steps and 2000 sampling steps. Convergence was assessed using the Rhat diagnostic [@stan]. Models were compared using approximate leave-one-out (LOO) cross-validation [@loo; @loo-paper] where negative values indicate an improved fit for the correlated model.


## Limitations


## Results 

### Summary

### Data description

```{r data, fig.cap = "Daily cases in England and by UKHSA region, with S-gene target result (failed, confirmed detected, or unknown), and centred 7-day moving average up to date of data truncation (dotted line). Source: UKHSA and coronavirus.gov.uk; data by specimen date."}
plot_daily_cases(
  daily, truncate_date = date_forecast_start, caption = "",
  start_date = date_data_start, smooth_total = TRUE
)
```


## References
